### 2024-07-05 rebuttals/rebuttals.tex

\documentclass[12pt,a4paper]{article}
\usepackage{longtable}
\usepackage[LGR,T1]{fontenc}
\usepackage{lscape}
%\usepackage{pdfsync}
\usepackage{multirow}
\usepackage{amsmath,bm} 
\usepackage{amsfonts}
\usepackage{amsthm} % Extended theorem environments
%\usepackage{amssymb} % Math symbols %%redundant with stix package
%\usepackage{esint} % Intégrales multiples
%\usepackage{esvect} % Vecteurs
\usepackage{mathtools}
\usepackage{pifont} 
\usepackage[version=4]{mhchem} % Chemical equations

\usepackage{fancyhdr}
\usepackage{graphicx}

\graphicspath{{../frontend/img/}{../chap_intro_ccl/img/}{../chap_case_study/img/}{../chap_ES_PCE/img/}{../chap_electro_uq/img/}{../chap_methodo/img/}{../chap_myopic/img/}{../chap_RL/img/}{../chap_atom_mol/img/}{../chap_RobPol/img/}{../appendices/img/}{../rebuttals/img/}} % Figures folder different for each input
%\graphicspath{{img/}} %
\DeclareGraphicsExtensions{.eps,.pdf,.png,.jpg}

@@

\usepackage{lastpage}
\usepackage{afterpage}
\usepackage{lettrine}
\usepackage{color,soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{colortbl}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{titlesec}
\def\eg{e.g.,\ }
\def\ie{i.e.,\ }

\newcommand{\clearemptydoublepage}{\newpage{\pagestyle{empty}\cleardoublepage}} %pour effacer les en-tetes sur la page vierge avant chaque chapitre

@@

%Palatino font
%\usepackage{pxfonts}
%\usepackage{libertine}
\usepackage[scaled=0.88]{beraserif}
\usepackage[scaled=0.85]{berasans}
\usepackage[scaled=0.84]{beramono}
\usepackage{mathpazo}
%\linespread{1.05}
\usepackage[T1,small,euler-digits]{eulervm}

\usepackage[nomessages]{fp}

\definecolor{bleuUCLclair}{rgb}{.09, 0.569, 1}
\definecolor{bleuUCLfonce}{rgb}{ .13, .52, .86}
\definecolor{redBurn}{rgb}{.91, 0.29, 0.08}

\usepackage{tocloft}

%\renewcommand{\cftsecpresnum}{Reviewer \#}
\renewcommand{\cftsecpresnum}{}
\renewcommand{\cftsecnumwidth}{6em}
\renewcommand{\cftsubsecpresnum}{}
\renewcommand{\cftsubsecnumwidth}{6em}

\setcounter{tocdepth}{3}
\setcounter{secnumdepth}{3}

\usepackage[nottoc]{tocbibind}
\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
\usepackage[square,numbers,sort&compress]{natbib}

@@

I would like to thank the jury members for the comments, which significantly helped improving the manuscript and substantiating the novelty of my work. Based on the notes taken during the private defense, I hereby transcribed, as accurately as possible, the jury's comments.

I believe I have addressed all the issues raised in the following answers. Some of them required adaptations of the text. These adaptations are either directly in the thesis manuscript or left for further developments in subsequent papers. For each comment, I have first highlighted the issue, then provided an answer, and finally described how the manuscript was adjusted, if needed.\\

When a comment explicitly comes from specific members of the jury, they are listed at the beginning of the comment with the following color code: {\color{orange} \textbf{Stefano} Moret}, {\color{teal} \textbf{Stefan} Pfenninger}, {\color{purple} \textbf{Sylvain} Quoilin} and {\color{violet} \textbf{Christophe} De Vleeschouwer}. Here is how an answer to a comment is structured:

\begin{mdframed}[style=comment] % Comment from the reviewer
Text of the comment. When a page number is mentioned at the beginning of the comment, it refers to the page number of the manuscript submitted before the private defense, which might differ from the one in its final version.
\end{mdframed}

@@

\begin{mdframed}[style=manuscript] % Comment from the reviewer

\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} - The variables and the uncertainties should be presented in a compact and mathematical way.
\end{mdframed}

\noindent In my opinion, the main variables of EnergyScope Pathway are presented {\color{blue} in Section 1.1 of the core of the manuscript}. Further details are also given {\color{blue}in Appendix B.2}. Finally, these variables are even exhaustively detailed in \cite{limpens2019energyscope} and \cite{limpens2024pathway} that are cited several times in the thesis manuscript.  Regarding the uncertainties, the way the ranges are defined and the way they affect the value of some parameters in the model are explained {\color{blue}in Section 1.2 of the core of the manuscript}. Consequently, no modification has been brought to the manuscript regarding this topic.

@@

\begin{mdframed}[style=manuscript] % Comment from the reviewer

\end{mdframed}

\subsection{Graphs and results}
\label{general_graphs_results}

\noindent To clarify this point, I have extended the caption of the related figures (see previous comments) and added two sentences in {\color{blue} Section 4.2.2}:

\begin{mdframed}[style=manuscript] % Comment from the reviewer
Since infeasible cases or those that overshoot the \ce{CO2}-budget are discarded before reaching 2050 (see Section 4.1.2), the number of attempts that reach further steps in the transition progressively decreases. Consequently, the share of successful transitions compared to failures progressively increases with time. 
\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Graphs like the ones presented during the private defense (slides 22 and 24) should be better explained. In general, improve the presentation of the results, especially for the RL chapter.
\end{mdframed}

@@

\noindent

\begin{mdframed}[style=manuscript] % Comment from the reviewer

\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}}, {\color{teal} \textbf{Stefan}} \& {\color{purple} \textbf{Sylvain}} - Many axes labels are missing, and the captions are often insufficient to read the figures. 
\end{mdframed}

\noindent I agree and have gone through all the figures of the manuscript to clarify them by adding axes labels or extending the caption where it was needed. Figures where y-axis (and its label) is missing are distribution (or density) plots (\ie Figures 3.6, 4.4, 4.5, 4.6, 4.8 and E.1).  Consequently, similarly to an histogram, the feature of interest and its span of values are represented on the x-axis. The y-axis has no real interest in terms of information. What matters is the shape of the curve to be able to point out for which values/ranges of values, the feature of interest occurred more or less often.  For this reason, I have not added y-axis for the aforementioned figures but rather improved their respective caption, emphasising, among other things, that these figures represent distribution of occurrences. For instance, the caption of {\color{blue}Figure 4.4} is now more explicit:

\begin{mdframed}[style=manuscript] % Comment from the reviewer
Exploration of the state space over the learning process: distribution of occurrence of cumulative emissions (left) and costs (right). Number of remaining attempts decrease with time since infeasible problem and solutions overshooting the \ce{CO2}-budget are discarded prematurely, \ie before reaching 2050. Distributions labelled as ``Failure'' represent the attempts that overshot the \ce{CO2}-budget by 2050 at the latest. The majority of successful transitions have cumulative emissions much lower than the \ce{CO2}-budget and are cheaper than the REF case. 
\end{mdframed}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{teal} \textbf{Stefan}} - Some big topics are missing like storage. Why is storage not more discussed? 
\end{mdframed}

\noindent As discussed during the private defense, storage is not more discussed in the manuscript for two reasons. First, the storage technologies, for the case of Belgium assuming a copper-plate approach, consist in two main technologies: methane storage mostly (about 15\,TWh) and, to a smaller extent, seasonal thermal storage (about 3\,TWh), with limited variations through the transition (see Figure \ref{fig:STORAGE_2}). Analyses under uncertainties have not shown major variations neither. Second, storage assets are intrinsically linked to the conversion technologies and, consequently, can be seen as by-products of the optimisation results. For these two reasons, I have rather focused my analyses on the use of resources and the installation and operation of conversion technologies. Given the discussion during the private defense and these further explanations, I have not brought modification to the manuscript in this regard.

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{STORAGE_2.pdf}
\caption{Storage installed capacities for the REF case}
\label{fig:STORAGE_2}
\end{figure}

@@

\end{mdframed}

\subsection{Structure}
\label{structure}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} \& {\color{teal} \textbf{Stefan}} - The thesis structure makes it quite difficult to follow. Having all the methodology summarised in Chapter 1, makes it difficult to link it to the different chapters. As changing the structure may be now a major work, I would invite you to think how to better connect the different parts. 
\end{mdframed}

\noindent Initially, I wanted to group methodology and results per kind of analysis, \ie UQ on Pathway (Chapter 3), RL (Chapter 4) and PCA (Chapter 5) but that would have made the RL, and even more PCA, methodology parts come too far in the manuscript. For this reason, I made the choice to group all the methodological approaches together in Chapter 1 and the results in their respective chapters.  Given this explanation, I have added one sentence {\color{blue}at the end of the introductory paragraph of the RL method in Chapter 1}:

\begin{mdframed}[style=manuscript] % Modification to manuscript
As Chapter 2 presents the input of the case study and Chapter 3 details the results of the GSA on the pathway model, the reader interested by the results of the RL method is invited to go to Chapter 4.
\end{mdframed}

\noindent and {\color{blue}at the end of the introductory paragraph of the PCA-based method in Chapter 1}:

\begin{mdframed}[style=manuscript] % Modification to manuscript
The reader interested by the results of the PCA-based method is invited to go to Chapter 5.
\end{mdframed}

\noindent Finally, I have added a schematic at the end of the Introduction of the thesis to clarify the structure and highlight that Chapter 4 and 5 can be read after the methodological sections 1.3 and 1.4, respectively (see Figure \ref{fig:intro:Thesis_Structure}).

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\textwidth]{Thesis_Structure.pdf}
\caption{Structure of the thesis. Beyond Chapter 2 that describes the case study of Belgium, Chapters 3, 4 and 5 collect the analyses resulting from the application of the different methodologies developed in Chapter 1. Given the (quasi-)independence of the analyses presented in the last three chapters, they can be read separately.}
\label{fig:intro:Thesis_Structure}
\end{figure}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} - Formal description of the problem is missing even in described in the text to make the link between the ideas and the implementation. Add some flow charts to introduce key elements
\end{mdframed}

\noindent This point is addressed with the structure chart presented in the previous comment as well as the improvement of Figure 1.4 presenting the implementation of the myopic approach as well as the way uncertainties and RL change the values of some parameters (see Figure \ref{fig:MY_process_code} of the present document).

@@

\subsection{References}
\label{references}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} \& {\color{purple} \textbf{Sylvain}} - I found it unusual to see many references to own work (Rixhon et al.) throughout the thesis. Unless the articles are unrelated to the thesis, these references should be removed and the associated content should be included in the manuscript. As an example, at p. 25, results are mentioned which are not reported in the thesis. The reader currently must refer to [48] to see these results, which should instead be included in the thesis. 
\end{mdframed}

\noindent This choice to sometimes refer to own works was made to focus only on the main contributions of the thesis. However, I agree that some content should be included in the thesis to enhance its reading. I hereby list the references to works I have been involved in (with the reference number of the manuscript version submitted before the private defense) and detail the potential modifications brought to the manuscript for each of them.

\begin{itemize}
\item \emph{[40] Limpens et al., Energyscope pathway: an open-source model to optimise the energy transition pathways of a regional whole-energy system}: Mostly cited to give credits to Gauthier Limpens' work in developing the Pathway version of EnergyScope, I have decided to refer to the paper about the formulation choices rather than including them in the thesis manuscript. Consequently, no further modification regarding this reference.
\item \emph{[48] Limpens et al., The impact of uncertainties on the Belgian energy system: Application of the Polynomial Chaos Expansion to the EnergyScope model}: The table gathering the results of the comparison between Morris' and PCE methods has been included {\color{blue}at the end of Section 1.2.2}:
\end{itemize}

\begin{mdframed}[style=manuscript] % Comment from the reviewer
In \cite{limpens2020impact}, we have assessed the PCE approach on the cost-optimum Belgian energy system in 2035, using EnergyScope TD. To do so, we compared the Top-14 most impacting parameters obtained from this approach with the one provided by the improved Morris method based on $\mu^*_{i}$. Even if the output of each method does not have the same physical meaning, both methods can rank the parameters by their impact on the total annual cost of the energy system. Both rankings were very similar which validated the use of PCE in the rest of this work (see Table \ref{tab:GSA:Comparison}).
\end{mdframed}

\begin{table}[htbp]
	\caption{Comparison of the Top-14 rankings for the improved Morris method (left) and total-order PCE method (right). We used the DTU's implementation of Morris method \cite{MorrisCodeDTU} (with $p=8$, $r=100$). }
    \label{tab:GSA:Comparison}
    \centering
    \begin{tabular}{lcc}
    \hline
		\textbf{Parameter} & \textbf{Morris Ranking} ($\mu^*_{i}$) & \textbf{PCE Ranking} ($S_i^T$) \\
        \hline
        Prices of hydrocarbons & 1 (0.7141) & 1 (0.6434) \\
        Cost of increased efficiency & 2 (0.3046) & 2 (0.1014) \\
        Cost of maintenance & 3 (0.2371) & 4 (0.0534) \\
        Discount rate & 4 (0.2019) & 5 (0.0467) \\
        Price of imported electricity & 5 (0.1974) & 3 (0.0809) \\
        CAPEX of the grid & 6 (0.1498) & 6 (0.0252) \\
        CAPEX of PV & 7 (0.1270) & 7 (0.0187) \\
        Increase of electricity demand & 8 (0.1168) & 8 (0.0154) \\
        CAPEX of nuclear power plant & 9 (0.0963) & 9 (0.0119) \\
        Increase of space heating demand & 10 (0.0856) & 11 (0.0088) \\
        Price of coal & 11 (0.0856) & 12 (0.0087)\\
        Price of renewable fuels & 12 (0.0794) & 10 (0.0104)\\
        Price of uranium & 13 (0.0607) & 13 (0.0047)\\
        Onshore wind load factor & 14 (0.0600) & 14 (0.0046)\\
       \hline
    \end{tabular}
\end{table}

\begin{itemize}
\item \emph{[49] Rixhon et al., The role of electrofuels under uncertainties for the Belgian energy transition}: This paper is cited to highlight the GSA performed on the snapshot model with different GWP limits. The main contribution of this work in the context of the thesis was to provide a preliminary screening and selection of parameters to carry out the GSA on the pathway model. The result of this selection is given in Section 2.4 where the uncertain parameters and their respective ranges are listed. Consequently, no further modification regarding this reference.
\item \emph{[104] Rixhon et al., Integration of non-energy among the end-use demands of bottom-up whole-energy system models}: Reference used to highlight the fact that whole-energy systems usually do not account for the non-energy demand. Consequently, no further modification regarding this reference.
\item \emph{[116] Rixhon et al., Comprehensive
integration of the non-energy demand within a whole-energy system: towards a defossilisation of the chemical industry in Belgium}: Reference to graphs that have either been adapted from the paper or directly taken from it. Consequently, no further modification regarding this reference.
\item \emph{[128] Rixhon et al., Taxonomy of the fuels in a whole-energy system}: Where the content of this paper is the Appendix A of the thesis, some references to this paper aimed at clarifying the meaning of ``renewable fuels'' (Section 2.2) or the fact that uranium is considered neither fossil nor renewable (Sections 3.1.2 and 4.2.2). Consequently, no further modification regarding this reference.
\end{itemize}

@@

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} - State of the art mainly European. Why not reference to other regions of the world.
\end{mdframed}

\noindent
It is an interesting topic for further investigations, especially how they address the subject in other parts of the world like Australia/Asia. This could be linked with the political perception of energy system optimisation models and the subsequent analyses. However, no modification has been brought to the manuscript regarding this topic.

\subsection{Writing style}
\label{writing_style}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{orange} \textbf{Stefano}} - While a more ``friendly'' writing style can be pleasant, there are many colloquial expressions that not do not fit well in the thesis, \eg ``carrot and stick'', ``apples with apples'', etc. I recommend reducing them to the bare minimum, or completely getting rid of them. 
\end{mdframed}

@@

\begin{mdframed}[style=manuscript] % Comment from the reviewer

\end{mdframed}

\begin{mdframed}[style=comment] % Comment from the reviewer
There are various typos in the thesis, e.g. ``storagR'' (page 3), ``therefore, therefore'' (p. 11), ``The'' (p.14), ``Moleclues'' (p. 61), etc.. Also, the use English language can be improved, e.g. ``calls FOR a variety'' (p. 9), ``cumulative emissions ARE'' (p. 14), ``than to'' (p. 34), etc. The document will benefit from a spelling/grammar check.
\end{mdframed}

\noindent The listed typos have been corrected and, overall, the manuscript has gone through Grammarly to improve the spelling and the grammar.

@@

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
In this sense, this work provided insight about possible transition pathways for Belgium to bring the technical dimension into the intrinsically political and interdisciplinary discussions and decisions that must be made in the coming years.
\end{mdframed}

\section{Chapter 1 - Methodology}
\label{methodo}

### 2024-07-05 appendices/app_EnergyScope.tex
Rebuttals: Keep going...

``\textit{Only when single-model results are contextualized by the model's position in the larger ensemble, the reader would be able to have a complete and correct interpretation of the output}" \cite{dekker2023identifying}. Energy system models of varying complexity are valuable tools for guiding policymakers and projecting future trends. These models enable the exploration of different energy scenarios and the assessment of their consequences. Specifically, techno-economic models play a crucial role in identifying technically feasible pathways for the energy transition while considering the associated economic costs. These models can be classified based on two key factors: technical resolution and simulation horizon, as illustrated in Figure \ref{fig:energy_models_classification}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{Models_classification.png}
\caption{Model can be classified by their core focus: \textbf{Operation} or \textbf{Design}. These categories can be broken down into subcategories. This work focuses on the system planning and scenario analysis models. Inspired from \cite{palmintier2013incorporating}. }
\label{fig:energy_models_classification}
\end{figure}

@@

These requirements are commonly found in reviews on energy system models. In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the ``popularity'' of the models via the number of downloads/sales or the integration of economic market equilibrium. Eight years later, \citet{lopion2018review} enriched the review of \citet{Connolly2010} using similar critieria and including models developed in the 2010s. In 2014, \citet{pfenninger2014energy} pointed out the current paradigms and challenges to face as well as the emerging approaches to address them in the 21st century energy systems modeling community.  Besides the behavioral and social factors, they also highlighted the challenges related to mutli-sectorial systems, time and space resolution or the open-accessibility of data and models and their ability to account for uncertainties. In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}" models, focusing on criteria (i-ii) and (iv). This review was followed by a classification where criteria (i-iv) were taken into account \cite{prina2020classification}. In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
Based on these reviews, models are compared based on all the previous criteria except the computational time (v) (see Table \ref{tab:intro_soa}). Indeed, the latter is hard to compare as models are not applied to the same case study and the information is rarely given. The table includes only the models that achieved partially at least four out of the five criteria. We endeavored to update the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.

@@

\noindent
Calliope is a ``\emph{tool that makes it easy to build energy system models}'' at different geographical scale. Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic. In fact, the model is typically employed for snapshot analysis, \ie optimization of a target future year. Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}. More recently, the model has been applied to analyse a scenario of a multi-energy district in Switzerland \cite{pickering2021quantifying}. Moreover, the model has been used with decades of weather data. However, its application has been limited to assessing the impact of inter-year variability in wind and PV on the results, 
rather than evaluating a transition pathway \cite{pfenninger2017dealing}. \\

@@

\noindent
Among the open-source models with an active community, PyPSA is one of the best-performing, with a large and active community, development at the state of the art, worldwide applications, and usage not only limited to academia. A study conducted by \citet{bartholdsen2019pathways} centered on Germany employed a representation comprising 16 time slices per representative year. This choice was substantiated by the work of \citet{welsch2014incorporating}, which demonstrated that this level of temporal granularity yields consistent results in comparison to hourly time resolution over a year. However, it is noteworthy that the utilization of a limited number of time slices may oversimplify the optimization of storage technologies, especially those designed for inter-month energy storage. This simplification can be viewed as a pragmatic approach to reduce the computational burden while over-simplifying the challenge of accurately integrating intermittent renewable energy sources. Furthermore, PyPSA, a modeling framework recognized for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}. \\

@@

\noindent
The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models. It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration. This model has a widespread adoption and has been utilized by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute. Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}. A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}. Recent developments by the IEA-ETSAP have resulted in a version that is compatible with the open-source solver CBC. In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilize 12 to 32 time-slices annually \cite{prina2020classification}. To highlight the sensitivity of results to time resolution, \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using up to 12 typical days with an hourly resolution. Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data is not ensured \cite{openmod_times_description}. \\

@@

\noindent
The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost and greenhouse gas emissions of the energy system. Typically, the two objectives are integrated by placing a limit on emissions while simultaneously minimizing the costs.

 \begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{chp_estd_overview.pdf}
\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimizing model (Optimisation) and results (Energy strategy). }
\label{fig:estd_overview}
\end{figure}

@@

In this section, we delve into the extension of EnergyScope TD from a static yearly snapshot model to a comprehensive pathway model. While snapshot models provide insights into the energy system for individual years, they lack the capacity to capture the dynamics inherent in investment strategies throughout a transition period. The proposed approach involves segmenting the transition into five-year intervals. This approach results in seven instances of EnergyScope TD -- called representative years -- spanning the 30-year transition period, covering the years from 2020 to 2050. To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition (see Figure \ref{fig:meth_path_methodology}). Overall, these constraints are integrated into a linear framework, ensuring computational efficiency, with an approximate computational time of 14 minutes on a personal laptop (2.4 GHz Intel Core i5 quad-core). Simplification and choices were necessary to implement linearly the problem while keeping a tractable computational time. In this section, the retained formulation is presented.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{ES_Pathway.pdf}
\caption{The pathway methodology relies on 7 representative years (blue boxes) where the model \gls{ESTD} is applied. Moreover, the formulation accounts for linking constraints (black boxes) and an initial condition (grey box). The overall problem is the pathway model.}
\label{fig:meth_path_methodology}
\end{figure}

@@

A part of the investment will remain after 2050. This residual investment, also called salvage value, can be calculated for each technology. 
A parameter, calculated \textit{a priori} , gives for each technology and construction phase, the remaining amount of years (\emph{remaining\_years}). 
As an example, if a PV panel has been built in 2045 and has a 20 years lifetime, the parameter will equal to 15 years. 
Thus, the salvage value is a fraction of the investment cost of this technology when it has been built. 
This fraction is the ratio between the number of remaining years and the lifetime of the technology. 
In the previous example, the residual investment of the PV built is 75\%. 
Equation (\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost. The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised. The residual value is not applied to technologies that are removed prematurely. This differ from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).

>>>>

``\textit{Only when single-model results are contextualised by the model's position in the larger ensemble, the reader would be able to have a complete and correct interpretation of the output}'' \cite{dekker2023identifying}. Energy system models of varying complexity are valuable tools for guiding policymakers and projecting future trends. These models enable the exploration of different energy scenarios and the assessment of their consequences. Specifically, techno-economic models play a crucial role in identifying technically feasible pathways for the energy transition while considering the associated economic costs. These models can be classified based on two key factors: technical resolution and simulation horizon, as illustrated in Figure \ref{fig:energy_models_classification}.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{Models_classification.png}
\caption{Models can be classified by their core focus: \textbf{Operation} or \textbf{Design}. These categories can be broken down into subcategories. This work focuses on the system planning and scenario analysis models. Inspired from \cite{palmintier2013incorporating}. }
\label{fig:energy_models_classification}
\end{figure}

@@

These requirements are commonly found in reviews on energy system models. In 2010, \citet{Connolly2010} reviewed 68 tools, considering similar criteria (\ie (i-iv) and (vi)), along with others such as the ``popularity'' of the models via the number of downloads/sales or the integration of economic market equilibrium. Eight years later, \citet{lopion2018review} enriched the review of \citet{Connolly2010} using similar critieria and including models developed in the 2010s. In 2014, \citet{pfenninger2014energy} pointed out the current paradigms and challenges to face as well as the emerging approaches to address them in the 21st century energy systems modeling community.  Besides the behavioral and social factors, they also highlighted the challenges related to mutli-sectorial systems, time and space resolution or the open-accessibility of data and models and their ability to account for uncertainties. In 2019, \citet{prina2019transition} reviewed 12 ``\emph{most established}'' models, focusing on criteria (i-ii) and (iv). This review was followed by a classification where criteria (i-iv) were taken into account \cite{prina2020classification}. In 2021, \citet{chang2021trends} conducted a survey-based review of 42 models for energy transition modelling, covering all criteria except computational time.
Based on these reviews, models are compared based on all the previous criteria except the computational time (v) (see Table \ref{tab:intro_soa}). Indeed, the latter is hard to compare as models are not applied to the same case study and the information is rarely given. The table includes only the models that achieved partially at least four out of the five criteria. We endeavored to update the model's information by consulting the model's website and repository, yet there is a possibility that some information might have been overlooked or omitted inadvertently.

@@

\noindent
Calliope is a ``\emph{tool that makes it easy to build energy system models}'' at different geographical scale. Even if the framework offers the possibility of modelling multi-year systems, we did not find a relevant publication on this topic. In fact, the model is typically employed for snapshot analysis, \ie optimisation of a target future year. Previous studies have used the model to investigate the phasing out of fossil and nuclear energies in a multi-regional UK power system \cite{pfenninger2015renewables}. More recently, the model has been applied to analyse a scenario of a multi-energy district in Switzerland \cite{pickering2021quantifying}. Moreover, the model has been used with decades of weather data. However, its application has been limited to assessing the impact of inter-year variability in wind and PV on the results, 
rather than evaluating a transition pathway \cite{pfenninger2017dealing}. \\

@@

\noindent
Among the open-source models with an active community, PyPSA is one of the best-performing, with a large and active community, development at the state of the art, worldwide applications, and usage not only limited to academia. A study conducted by \citet{bartholdsen2019pathways} centered on Germany employed a representation comprising 16 time slices per representative year. This choice was substantiated by the work of \citet{welsch2014incorporating}, which demonstrated that this level of temporal granularity yields consistent results in comparison to hourly time resolution over a year. However, it is noteworthy that the utilisation of a limited number of time slices may oversimplify the optimisation of storage technologies, especially those designed for inter-month energy storage. This simplification can be viewed as a pragmatic approach to reduce the computational burden while over-simplifying the challenge of accurately integrating intermittent renewable energy sources. Furthermore, PyPSA, a modeling framework recognised for its robustness and active user community, has also been employed to investigate scenarios related to myopic transitions \cite{pedersen2022long}. \\

@@

\noindent
The TIMES model, short for The Integrated MARKAL-EFOM System, is a well-established framework renowned for its capacity to generate comprehensive energy models. It encompasses a rich array of features, including support for multi-cell modeling, pathway analysis, full-scale representation of energy systems, and the consideration of market equilibrium dynamics, all of which facilitate thorough scenario exploration. This model has a widespread adoption and has been utilised by worldwide institutions such as the \gls{IEA} or technical ones such as VITO (Vlaamse Instelling voor Technologisch Onderzoek) research institute. Notably, TIMES was reported as commercial (\ie not free to download) in 2010 \cite{Connolly2010}. A more recent survey conducted in 2020-2021 confirmed that the model was using a commercial interface \cite{chang2021trends}. Recent developments by the IEA-ETSAP have resulted in a version that is compatible with the open-source solver CBC. In various studies conducted in different regions, including Canada, Sweden, the EU, and Denmark, TIMES has been shown to utilise 12 to 32 time-slices annually \cite{prina2020classification}. To highlight the sensitivity of results to time resolution, \citet{haydt2011relevance} conducted a study focusing on the electrical sector, using up to 12 typical days with an hourly resolution. Regarding data accessibility, while some publications partially present the used dataset, the overall accessibility of TIMES data is not ensured \cite{openmod_times_description}. \\

@@

\noindent
The conceptual structure of the model is illustrated in Figure \ref{fig:estd_overview}: given the end-use energy demand, the efficiency and cost of energy conversion technologies, the availability and cost of energy resources, the model identifies the optimal investment and hourly operation strategies to meet the demand and minimise the total annual cost and greenhouse gas emissions of the energy system. Typically, the two objectives are integrated by placing a limit on emissions while simultaneously minimising the costs.

 \begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{chp_estd_overview.pdf}
\caption{EnergyScope TD model is a flow model with inputs (Parameters), an optimising model (Optimisation) and results (Energy strategy). }
\label{fig:estd_overview}
\end{figure}

@@

In this section, we delve into the extension of EnergyScope TD from a static yearly snapshot model to a comprehensive pathway model. While snapshot models provide insights into the energy system for individual years, they lack the capacity to capture the dynamics inherent in investment strategies throughout a transition period. The proposed approach involves segmenting the transition into five-year intervals. This approach results in seven instances of EnergyScope TD -- called representative years -- spanning the 30-year transition period, covering the years from 2020 to 2050. To bridge these representative years, we introduce additional constraints that capture the investments changes between consecutive periods, accounting for societal inertia and evaluating both the cost implications and emissions of the transition (see Figure \ref{fig:meth_path_methodology_app}). Overall, these constraints are integrated into a linear framework, ensuring computational efficiency, with an approximate computational time of 14 minutes on a personal laptop (2.4 GHz Intel Core i5 quad-core). Simplification and choices were necessary to implement linearly the problem while keeping a tractable computational time. In this section, the retained formulation is presented.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{ES_Pathway.pdf}
\caption{The pathway methodology relies on 7 representative years (blue boxes) where the model \gls{ESTD} is applied. Moreover, the formulation accounts for linking constraints (black boxes) and an initial condition (grey box). The overall problem is the pathway model.}
\label{fig:meth_path_methodology_app}
\end{figure}

@@

A part of the investment will remain after 2050. This residual investment, also called salvage value, can be calculated for each technology. 
A parameter, calculated \textit{a priori} , gives for each technology and construction phase, the remaining amount of years (\emph{remaining\_years}). 
As an example, if a PV panel has been built in 2045 and has a 20 years lifetime, the parameter will equal to 15 years. 
Thus, the salvage value is a fraction of the investment cost of this technology when it has been built. 
This fraction is the ratio between the number of remaining years and the lifetime of the technology. 
In the previous example, the residual investment of the PV built is 75\%. 
Equation (\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost. The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised. The residual value is not applied to technologies that are removed prematurely. This is similar to the work of \citet{prina2019transition} but differs from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).

In the work of \citet{goffauxpathway}, we carried out a sensitivity analysis on the formulation of the salvage value. First and foremost, this analysis confirmed the need to account for the salvage value to avoid penalising the capital intensive technologies towards the end of the transition \cite{poncelet2016myopic}. Then, the sensitivity analysis focused on two elements in the expression of the salvage value: $\tau_{phase}(p)$ and $\textbf{F\textsubscript{decom}}(p2,p,i)$.

In Equation (eq:salvage_app), $\tau_{phase}$ is the annualisation factor corresponding to the phase where the technology was built. With this expression, if a technology is still operational for half its lifetime after 2050, then half its initial investment is subtracted to the total transition cost. However, one could wonder if a technology that is still operational for half of its lifetime is not worth less than half of its initial investment cost, given the depreciation of the asset. In that case, the annualisation factor should account for the last phase of the optimisation (\ie 2045\_2050) when the residual value of the technologies are calculated and subtracted from the total transition cost.

On top of this, we also investigated the impact of accounting for the decommissioned technologies or not in the expression of the salvage value.  In Equation (eq:salvage_app), if a technology is decommissioned before 2050, then the salvage value of this technology will be 0. This is relevant since a technology decommissioned before 2050 would not be available after 2050. The issue is that the model could keep unnecessary technologies to subtract their salvage value from the total transition cost in order to decrease it. This would be done if the fixed OPEX of an unused technology is lower than its salvage value. Therefore, one could wonder if it would not be better to take into account the decommissioned technologies in the salvage value. In that case, a technology that would have been operational after 2050 but that had been prematurely decommissioned before 2050, would have a salvage value.

The overall conclusion of this sensitivity analysis was that the salvage value expression has little influence on the energy system at the end of the time horizon.
